\section{Report Outline}

\newcommand{\outlineref}[1]{\autoref{#1} \nameref{#1}}

\paragraph{\outlineref{chap:intro}} introduces and gives motivation for using {\rust} in an embedded system.
The interpretation of the assignment and an outline of the project is presented along with a summary of the contributions of the project.

\paragraph{\outlineref{chap:background}} provides background material for the rest of the thesis.
The Rust programming language is introduced along with the bundled package manager, {\cargo}.
Further, the existing hardware platform, EFM32, and software libraries used for developing embedded applications are presented.

%\paragraph{\outlineref{chap:impl}} looks at the components developed to provide the {\rust} platform.
%As most of these components are largely orthogonal to each other, each are described in a separate section followed by a small discussion, where applicable.
%In addition, the two projects developed to evaluate the platform in \autoref{chap:results} is described.

\paragraph{\outlineref{chap:startup}} presents the what happens in order for a {\rust} program to start executing on the \gls{mcu}.

\paragraph{\outlineref{chap:rel}} gives an overview over the {\rust} standard library for embedded systems.

\paragraph{\outlineref{chap:bindings}} goes in to detail on the bindings developed for the peripheral libraries used to control the \gls{mcu}.

\paragraph{\outlineref{chap:build}} looks at the build system used to build application for the {\rg} platform.

\paragraph{\outlineref{chap:app}} presents some high level libraries and some application build on the core {\rg} functionality.

\paragraph{\outlineref{chap:results}} presents how the platform was evaluated and the results from the evaluation.
The platform was evaluated by considering \emph{code size}, \emph{performance} and \emph{energy efficiency}.

\paragraph{\outlineref{chap:discussion}} provides a discussion of the platform and the results presented in \autoref{chap:results}.

\paragraph{\outlineref{chap:conclusion}} presents a conclusion based on the state of the platform and the results.

\paragraph{\outlineref{chap:future}} outlines possible extensions and suggests further work based on this project.
