% !TEX root = ../main.tex

\subsection{EFM32}
\label{sub:emf32}

EFM32 refers to a family of 32-bit ARM microcontrollers that is developed by Silicon Labs. They are
based on the ARM Cortex-M series of RISC processor cores, a range of different processor cores
intended for microcrontrollers, that focuses on low cost and energy-usage. These factors are crucial
in modern systems and applications where energy efficiency is of great importance. For example with
the \gls{iot}, where we predict a future where tens of billions of devices will be connected to the
Internet, ranging from Super Computers down to small embedded devices that might be used to power up
and control everything from cars to light bulbs via the Internet. The different processor cores of
the Cortex-M family are summarized in \autoref{tab:cortex_m}, ranging from the smallest and simplest
processor cores, to the more advanced ones with more features.

\begin{centering}
\begin{tabular}{c|c}
Name & Purpose/features
Cortex-M0 & Lowest cost and lowest area
Cortex-M0+ & Lowest power
Cortex-M1 & Designed for implementation in FPGAs
Cortex-M3 & Performance efficiency
Cortex-M4 & DSP, SIMD, FP
Cortex-M7 & Cache, TCM, AXI, ECC, double + single FP
\caption{Cortex-M}
\label{tab:cortex_m}
\end{tabular}
\end{centering}

The EFM32 family of microcontrollers are all based on different Cortex-M processors, some of their
features are summarized in \autoref{tab:efm32_family}. The focus of these microcontrollers is energy
efficiency and low power-consumption in resource-constraint environments. The microcontrollers
implement several different methods for reducing the power consumption. The most important way to
achieve low power consumption is by turning off the different part of the processor that are
inactive, so that these parts do not longer draw any power from the overall system. The EFM32
processors features five different energy modes, or sleep modes, ranging from EM0 (Energy Mode 0) to
EM4, where EM4 is the lowest and wakeup on interrupts. The different peripherals provided with the
EFM32's operate in different energy modes. This allows applications to utilize many different
peripherals for data collection and communication while the processor itself is turned off, but with
the opportunity to wake up the processor on different interrupt-signals in order to do general
processing.

\begin{centering}
\begin{tabular}{c|c|c}
Name & Processor & Speed (MHz)
Zero Gecko    & ARM Cortex-M0+ & 24
Tiny Gecko    & ARM Cortex-M3  & 32
Gecko         & ARM Cortex-M3  & 32
Leopard Gecko & ARM Cortex-M3  & 48
Giant Gecko   & ARM Cortex-M3  & 48
Wonder Gecko  & ARM Cortex-M4  & 48
\caption{EFM32}
\label{tab:efm32_family}
\end{tabular}
\end{centering}



\subsection{Emlib} % (fold)
\label{sub:emlib}

\emlib is a peripheral \gls{api} by Silicon Labs that provides a nice abstraction
layer for their EFM32 series of ARM micro-controllers. It is a general library written in C that
provides functions and modules to read, write and control the different peripherals that is
supported by the EMF32 micro-controllers. With the \gls{api}, the application programmer can control
and set up the different peripherals and easily control sleep modes and wakeup, and interrupt
handlers among others. The library is built on the Cortex Microcontroller Software Interface Standard
(CMSIS) provided by ARM. It is possible to control the EFM32's with these libraries as well, but
\emlib is designed to as a standalone library.
